\emph{Gaia} DR2 adds clues for dynamical ejection of a $\sim$$150\,M_\odot$ star

### 2018-05-23 VFTS682.tex
major, including danny's fig

% Recently, \citet{platais:18} used HST observations (based on GO-12499
% \& 13359 observations with WFC3/UVIS and on archival WFPC2 and
% GO-12499 with both cameras, see \citealt{platais:15}) to determine the
% proper motion of massive stars in the 30 Doradus region. Our star of
% interest is not saturated in any HST image \todo{?},
% because of the large extinction toward this star.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS 682 relative to the field, providing a
measure of the motion of this star completely independent from
ours. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$.

% Since these measures are completely independent from ours, we can take
% an average weighted with $1/\delta \mu_i^2$ (cf.~notation in
% \Eqref{eq:mean}), obtaining $\delta\mu_\mathrm{RA}=\pm\,\mathrm{mas\
%   yr^{-1}}$ and $\delta\mu_\mathrm{DEC}=\pm\,\mathrm{mas\
%   yr^{-1}}$.

% %% pure HST physical velocities
% \todo{done in the train: to be checked} Converting these into physical velocities we obtain 
% $\delta v_\mathrm{RA}=2.3\pm30\,\kms$, $\delta v_\mathrm{DEC}=47\pm\,23\kms$, corresponding to
% $v_\parallel^\mathrm{HST} = 47\pm\,1\kms$.

\todo{redo below with HST only and Gaia only?Redo also calculation
  of $v_\mathrm{pec}$}

\subsection{New data from \emph{Gaia} DR2  \label{data:gaia}}
%% using
%% /scratch/mathieu/MMPS/SCRIPTS_dirty/my_script/binaries/binary_c/30Dor/vfts682/data/asu.tsv to find average

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS 682. }
  \begin{tabular}{l|c|c}
  \tableline
    \tableline
    Parameter & Value & Source\\
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                      & \multirow{4}{*}{\emph{Gaia} DR2}\\[5pt]
    DEC \hfill [degree] & -69.07 % $\pm$  0.05
                      & \\[5pt]
    \cline{1-2}
    $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.12$\pm$0.07 & \\[5pt]
    $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.08 & \\[5pt]
    \hline
    $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
    $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                      0.2$\pm$0.1 &
    \\[5pt]
    \hline
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
    \tableline
  \end{tabular}
  \tablecomments{We do not list the error on the RA and DEC positions,
    which can be retrieved from the Gaia DR2 catalog and are of order
    $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$. We adopt the peculiar
    radial velocity calculated using the  NV $\lambda4944$ line, which is
    less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).}
    \end{center}
  \label{tab:vfts682}
\end{table}

@@

% Using the stars selected as described in \Secref{data:gaia} (a
% subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
% $\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
% $\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
%   yr^{-1}}$. These values are
% in good agreement with what is found by \cite{lennon:18}. However, we
% emphasize their sensitivity to the choice of stars adopted to define
% the reference frame.
Subtracting the mean proper motion components given by
\citep{lennon:18} from the
proper motion of VFTS 682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA}^{Gaia} = \pm \,\mathrm{mas\ yr^{-1}}$ and $\delta\mu_\mathrm{DEC}^{Gaia} =
\pm \,\mathrm{mas\ yr^{-1}}$. The error budget is
dominated by the errors on the proper motion components of VFTS 682.

These can be converted into
the components of the relative transverse velocity $\delta v_\mathrm{RA}^{Gaia}=28\pm16\,\kms$,
$\delta v_\mathrm{DEC}^{Gaia}=20\pm19\,\kms$, assuming a distance of
50\,kpc. % (we do not account for the uncertainty in the distance
% estimate when propagating errors).
The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the three-dimensional
peculiar speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
    +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
      v_\mathrm{rad}^{VFTS}\right)^2} = 46 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS 682 with respect to
the surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

>>>>

The catalog from \citet{platais:18} lists the proper motion
components of VFTS 682 relative to the field, providing a
measure of the motion of this star completely independent from
ours. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$, corresponding to a relative proper motion

\begin{equation}
  \label{eq:pm_around}
  \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
  = 0.20 \pm 0.15 \ \ .
\end{equation}

These proper motion components can also be converted into projected
velocities assuming a distance of 50\,kpc, obtaining $\delta
v_\mathrm{RA}^{HST}=2\pm31\kms$ and $\delta
v_\mathrm{RA}^{HST}=47\pm24\kms$, corresponding to a two-dimensional
projected velocity of $47\pm24\kms$. Including the radial velocity
measured by the VFTS survey we obtain a three-dimensional velocity of

\begin{equation}
  \label{eq:speed_around_HST}
  v_\mathrm{pec}^{HST} = \sqrt{\left(\delta v_\mathrm{RA}^{HST}\right)^2
    +\left(\delta v_\mathrm{DEC}^{HST}\right)^2+\left(\delta
      v_\mathrm{rad}^{VFTS}\right)^2} = 56 \pm 23
  \kms \ .
\end{equation}

\subsection{New data from \emph{Gaia} DR2  \label{data:gaia}}

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS 682. }
  \begin{tabular}{l|c|c}
  \tableline
    \tableline
    Parameter & Value & Source\\
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                      & \multirow{4}{*}{\emph{Gaia} DR2}\\[5pt]
    DEC \hfill [degree] & -69.07 % $\pm$  0.05
                      & \\[5pt]
    \cline{1-2}
    $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.10$\pm$0.07 & \\[5pt]
    $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.09 & \\[5pt]
    \hline
    $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
    $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                      0.2$\pm$0.1 &
    \\[5pt]
    \hline
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
    \tableline
  \end{tabular}
  \tablecomments{We do not list the error on the RA and DEC positions,
    which can be retrieved from the Gaia DR2 catalog and are of order
    $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$. We adopt the peculiar
    radial velocity calculated using the  NV $\lambda4944$ line, which is
    less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).}
    \end{center}
  \label{tab:vfts682}
\end{table}

@@

% Using the stars selected as described in \Secref{data:gaia} (a
% subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
% $\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
% $\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
%   yr^{-1}}$. These values are
% in good agreement with what is found by \cite{lennon:18}. However, we
% emphasize their sensitivity to the choice of stars adopted to define
% the reference frame.
Subtracting the mean proper motion components given by
\citep{lennon:18} from the
proper motion of VFTS 682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA}^{Gaia} = 0.10 \pm 0.07\,\mathrm{mas\ yr^{-1}}$
and $\delta\mu_\mathrm{DEC}^{Gaia} = 0.08
\pm 0.09\,\mathrm{mas\ yr^{-1}}$. These components result in a
two-dimensional relative proper motion of
\begin{equation}
  \label{eq:pm_gaia_around}
  \delta \mu^{Gaia} = \sqrt{\left(\delta\mu_\mathrm{RA}^{Gaia}\right)^2+\left(\delta\mu_\mathrm{DEC}^{Gaia}\right)^2}
  = 0.13\pm 0.09 \ \ .
\end{equation}

Figure \Figref{fig:pm_polar} shows the distribution in relative proper motion
of the stars used in \cite{lennon:18} to define the frame of reference
we also adopt here, cf.~the right panel in their Fig.~2. The
proper motions are decomposed in the radial and tangential direction
from R136, to highlight the likelihood of it being the origin of fast
moving stars. The plus sign marks VFTS 682, which sits at the edge of
the proper motion distribution, but is not a clear outlier. Most
relevant is the direction of the proper motion of VFTS 682, which we
discuss in \Secref{sec:r136_origin}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/figure_polar_682-1.pdf}
  \caption{Polar plot of the relative proper motion components for the stars
    defining our reference frame. VFTS682 is indicated by the plus sign, and
    other notable outliers (see \citet{lennon:18}) are also labeled. Concentric dashed-lines denote relative proper motions of
    0.1, 0.2, and 0.4$\,\mathrm{mas\ yr^{-1}}$. Positive angle
    indicate a tangential component pointing in the counterclockwise
    direction, with 0 degrees corresponding to proper motion pointing
    radially outward from R136.
}
  \label{fig:pm_polar}
\end{figure}

These can be converted into
the components of the relative transverse velocity $\delta v_\mathrm{RA}^{Gaia}=24\pm19\,\kms$,
$\delta v_\mathrm{DEC}^{Gaia}=20\pm23\,\kms$, assuming a distance of
50\,kpc. These can be converted in a projected two-dimensional
velocity of $31\pm21\kms$. % (we do not account for the uncertainty in the distance
% estimate when propagating errors).
The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the three-dimensional
peculiar speed of the star:

\begin{equation}
  \label{eq:speed_around_Gaia}
  v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
    +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
      v_\mathrm{rad}^{VFTS}\right)^2} = 44 \pm 21
  \kms \ .
\end{equation}

Therefore, two completely independent measures of the proper motion of
VFTS 682 relative to the surrounding field, one from HST and one from
\emph{Gaia} DR2, yield values of the peculiar three-dimensional
velocity of VFTS 682 (\Eqref{eq:speed_around_HST} and
\Eqref{eq:speed_around_Gaia}, respectively) which would make it the most massive runaway star
known to date. However, the large errors on both the proper motion measures
require confirmation with future astrometric data.
